Leadership across cultures and contexts: an exploratory analysis

### 2020-05-13 leadership_across_cultures_contexts.Rmd
Edits in conclusion.

# Concluding remarks: Beyond the non-WEIRD 
<!-- NOT SURE OF TITLE -- WANT TO EMPHASIZE NON WEIRD DIVERSITY AND ALSO MENTION BACK TO WERID FINDINGS BRIEFLY -->
In recognizing the limitated generalizability of most extant psychological, sociological, and managerial data, many social scientists are now diversifying their samples to better understand *human* psychology, behavior, and culture. We caution that the WEIRD people problem in behavioral science cannot be remedied simply by a "non-WEIRD" solution. Our results, and much of anthropology and ethnography, clearly demonstrate enormous diversity within populations that are now often categorized as non-WEIRD. Furthermore, higher-order explanatory variables such as "subsistence strategy" are probably less relevant for understanding the complexity of behavioral diversity than more salient group contexts. Despite limitations, the ethnographic record presents an invaluable opportunity to explore patterns of human cultural and behavioral diversity. Evolutionary theorists and researchers looking beyond WEIRD samples should take special interest in comparative ethnographic methods.

@@

While there is cross-cultural evidence of a dominance style of leadership involving coercive authority, confidence and aggression, as well as a prestige style involving special knowledge, experience, and innovation, we suggest rethinking dichotomous models of leadership styles and focusing on inter-individual variation in specific leadership competencies. There is likely to be substantial overlap in and interaction between physical, spiritual, cognitive, and other prosocial competencies and the degree to which a leader relies on such competencies will be shaped by their individual levels in them, the competencies of potential rivals, as well as their functions and group context, follower preferences, and cultural norms. For example, widespread preferences for egalitarianism and social equality are likely strong drivers of followers preferences for leaders across diverse societies [e.g., @cheng_dominance_2019; @price_evolution_2014], but shamanism and ritual based contexts represent a social domain in which ideologies support a style of leadership fusing both dominance and prestige-based competencies. It seems evident that cultural forces and ideologies play important roles in shaping leadership styles and follower preferences including among WEIRD populations and in national political movements around the world [e.g., @mcdermott_leadership_2018; @inglehart_trump_2016].

One interesting pattern that has not been widely discussed in the evolutionary literature on leadership is the role of shamans and their intersection, in our results, with the dominance and prestige styles of leadership. Our competencies cluster (Figure \@ref(fig:cluster)A) includes the *Supernatural* abilities dimension. @winkelman1987cross found that in 47% of societies sampled from the HRAF ($n = 43$) shamans were associated with providing decision-making services, a proscial and prestige-based computational service, and associated with military power, which is often associated with dominance. Providing ritual functions and military leadership are common (documented in over 50% of societies here) functions leaders serve (see Figure \@ref(fig:varsupport)).

@@

Our results also suggest both broad continutites and contrasts between leadership among the nonindustrial, non-WEIRD populations sampled here and the WEIRD samples common in managerial studies. Our organize and strategize features of leader functions reflect the @wofford_factor_1970 *order and group achievement* and *personal interaction* factors, respectively. Also, Woffords *personal enhancement* and *dynamic and achievement* factors are similar to leadership dimensions captured by our compentencies feature. Our results, however, emphasize the prosociality of leaders in non-WEIRD settings over an emphasis on self-presevation, relative to WEIRD leaders (e.g., Woffords *security and maintenace* factor). We also identified conformity to social norms as a feature of human leadership across cultures, similar to predictions and findings drawing on Culturally Endorsed Implicit Leadership Theory by managerial scholars working with WEIRD populations [@behrendt_integrative_2017]. Our results on the functions of leaders across cultures and contexts also corresponds to the *task-oriented* vs. *relation-oriened* framework of managerial beahviors (i.e., our Managemebt vs. Prosociality features of leader functions), however, our results do not support the strong emphasis on charismastic leadership found by the GLOBE CEO study [@house2013strategic] (although charism was documented as a quality of leaders in over 20% of cultures and oratory skills in over 50%). See Figure \@ref(fig:cluster).

These exploratory results provide an empirical foundation for future research on leadership across or within diverse cultural settings and levels of social organization. Political leadership is common across a broad range of populations, including among relatively small-scale and egalitarian hunter-gatherers. Among hunter-gatherers, however, leaders more frequently emerge within residential subgroups and kin groups. Given the vast majority of human evolutionary history involved subsistence economies characterized by hunting and gathering and greater residential mobility [@lee_cambridge_1999; @marlowe_hunter-gatherers_2005], these results suggest leadership within kin groups and local subgroups likely has a long evolutionary history. Despite a universal male-bias in political leadership at the community level, women are expected to have greater decision-making responsibilities and adopt leadership roles within families and local groups [@brown_her_1985; @garfield_evolutionary_2019; @hagen_leadership_2019; @smith_obstacles_2018; @von_rueden_sex_2018]. Given the broad range of leadership dimensions represented in the ethnographgric record, we suggest leaders-follower dynamics should be increasingly incorporated in evolutionary models beahvior and group dynamics, including demography, social complexity and inequality, cooperation and punishment, and resource exchange [c.f., @bird_variability_2019; @ember_our_2018; @micheletti_modelling_2020; @smith_cultural_2020; @birch2017philosophy; @marlowe_more_2008; @henrich_markets_2010; @mace_co-evolution_1998; @mace_phylogenetic_2005; @mattison_evolution_2016].

@@

```{r sexsubtable}
kable(sub_sex_tbl, 
      caption = 'Distributuon of text records by leader sex and subsistence strategy.')
```

>>>>

# Concluding remarks: Beyond the non-WEIRD 
<!-- NOT SURE OF TITLE -- WANT TO EMPHASIZE NON WEIRD DIVERSITY AND ALSO MENTION BACK TO WERID FINDINGS BRIEFLY -->
In recognizing the limited generalizability of most extant psychological, sociological, and managerial data [@henrich_weirdest_2010], many social scientists are now diversifying their samples to better understand *human* psychology, behavior, organizations, and culture. We caution that the WEIRD people problem in behavioral science cannot be remedied simply by a "non-WEIRD" solution. Our results, and much of anthropology and ethnography, clearly demonstrate enormous diversity within populations that are now often categorized as "non-WEIRD", "traditional", or "small-scale" [e.g., @reyes-garcia_small-scale_2017]. Furthermore, higher-order explanatory variables such as "subsistence strategy" are probably less relevant for understanding the complexity of behavioral diversity than more salient group contexts. Despite limitations, the ethnographic record presents an invaluable opportunity to explore patterns of human cultural and behavioral diversity. Evolutionary theorists and researchers looking beyond WEIRD samples should take special interest in comparative ethnographic methods.

@@

Cross-cultural evidence supports a dominance style of leadership involving coercive authority, confidence and aggression, as well as a prestige style involving special knowledge, experience, and innovation. We suggest, however, moving beyond dichotomous models of leadership strategies and focusing on inter-individual variation in broader typologies of leadership competencies. There is likely to be substantial overlap in, and interaction between physical, supernatural, cognitive, and other prosocial competencies. The degree to which a leader relies on such competencies will be shaped by their individual levels of them, the competencies of potential rivals, as well as their functions and group context, follower preferences, and cultural norms. For example, widespread preferences for egalitarianism and social equality are likely strong drivers of followers preferences for leaders across diverse societies [e.g., @cheng_dominance_2019; @price_evolution_2014], but shamanism and ritual based contexts represent a social domain in which ideologies support a style of leadership fusing both dominance and prestige-based competencies. Cultural forces and ideologies play important roles in shaping leadership styles and follower preferences including among WEIRD populations and in national political movements around the world [e.g., @mcdermott_leadership_2018; @inglehart_trump_2016] and the ecology of leadership dimensions can be better captured than dichotomous frameworks offer.

One interesting pattern that has not been widely discussed in the evolutionary literature on leadership is the role of shamans and their intersection, in our results, with the dominance and prestige styles of leadership. Our competencies cluster (Figure \@ref(fig:cluster)A) includes the *Supernatural* qualities dimension. @winkelman1987cross found that in societies sampled from the HRAF ($n = 43$) shamans provided decision-making services within judiciary, economic, military, or political domains in 44% or more of societies, and held charismatic leadership roles in 19% of societies; @winkelman1992shamans emphasizes the influential positions of leadership shamans attain through their charisma, social unification, healing abilities, and use of supernatural powers to cause harm. Providing ritual functions and military leadership are common (documented in over 50% of societies here) functions leaders serve (see Figure \@ref(fig:varsupport)).

@@

Our results also suggest both broad continuities and contrasts between leadership among the nonindustrial, non-WEIRD populations sampled here and the WEIRD samples common in managerial studies. Our organize and strategize features of leader functions reflect the @wofford_factor_1970 *order and group achievement* and *personal interaction* factors, respectively. Also, Woffords *personal enhancement* and *dynamic and achievement* factors are similar to leadership dimensions captured by our competencies feature. Our results, however, emphasize the prosociality of leaders in non-WEIRD settings over an emphasis on self-preservation, relative to WEIRD leaders (e.g., Woffords *security and maintenance* factor). We also identified conformity to social norms as a feature of human leadership across cultures, similar to predictions and findings drawing on Culturally Endorsed Implicit Leadership Theory by managerial scholars working with WEIRD populations [@behrendt_integrative_2017]. Our results on the functions of leaders across cultures and contexts also corresponds to the *task-oriented* vs. *relation-oriented* framework of managerial behaviors (i.e., our Management vs. Prosociality features of leader functions), however, our results do not support the strong emphasis on charismatic leadership found by the GLOBE CEO study [@house2013strategic] (although charisma was documented as a quality of leaders in over 20% of cultures and oratory skills in over 50%). See Figure \@ref(fig:cluster).

These exploratory results provide an empirical foundation for future research on leadership across or within diverse cultural settings and levels of social organization. Political leadership is common across a broad range of populations, including among relatively small-scale and egalitarian hunter-gatherers. Among hunter-gatherers, however, leaders more frequently emerge within residential subgroups and kin groups. Given the vast majority of human evolutionary history involved subsistence economies characterized by hunting and gathering and greater residential mobility [@lee_cambridge_1999; @marlowe_hunter-gatherers_2005], these results suggest leadership within kin groups and local subgroups likely has a long evolutionary history. Despite a universal male-bias in political leadership at the community level, women are expected to have greater decision-making responsibilities and adopt leadership roles within families and local groups [@brown_her_1985; @garfield_evolutionary_2019; @hagen_leadership_2019; @smith_obstacles_2018; @von_rueden_sex_2018]. Given the broad range of leadership dimensions represented in the ethnographic record, we suggest leaders-follower dynamics should be increasingly incorporated in evolutionary models behavior and group dynamics, including demography, social complexity and inequality, cooperation and punishment, and resource exchange [c.f., @bird_variability_2019; @ember_our_2018; @micheletti_modelling_2020; @smith_cultural_2020; @birch2017philosophy; @marlowe_more_2008; @henrich_markets_2010; @mace_co-evolution_1998; @mace_phylogenetic_2005; @mattison_evolution_2016].

@@

```{r sexsubtable}
kable(sub_sex_tbl, 
      caption = 'Distribution of text records by leader sex and subsistence strategy.')
```
